---
title: "Final_project"
author: "Cade Garcia"
date: "2025-08-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Tsunami Risk Map

Following the 8.8 magnitutde earthquake, largest ever recorded earthquake in Russia since 1952. Hawai'i along with many other states and some countries were given a tsunami warning. This threw our state, particularly the island of O'ahu into chaos. Along with the regular traffic, there were floods of panicking people trying to evacuate causing an overwhelming build up of traffic on freeways and gridlock in the cities. This has revealed two things. One, that many of us lack preparation for natural disasters and two, we need to be more self sufficient and independent. Part of the problem is that we relied too heavily on the government to save us/ be prepared for us.

The purpose of this map is to bring attention to the need for disaster preparedness especially to those that are in high risk zones.

#### **Sources**

*The two geojson files are available in github.*

-   API from US census bureau from 2024

-   TSUNAMI ZONES :

    -   [https://opendata.hawaii.gov/dataset/tsunami-evacuation-all-zones/resource/36d92c5f-c0f7-4a96-9f59-0a0f153c2c17?inner_span=Truehttps://opendata.hawaii.gov/dataset/tsunami-evacuation-all-zones/resource/36d92c5f-c0f7-4a96-9f59-0a0f153c2c17?inner_span=True](https://opendata.hawaii.gov/dataset/tsunami-evacuation-all-zones/resource/36d92c5f-c0f7-4a96-9f59-0a0f153c2c17?inner_span=True){.uri}

-   TRAFFIC :

    -   <https://opendata.hawaii.gov/dataset/traffic-analysis-zones-taz-oahu?utm_source=chatgpt.com>

```{r}
library(shiny)
library(mapgl)
library(mapboxapi)
library(sf)
library(tidycensus)
```

### Loading Data

```{r}
# Get population for Hawaii census tracts
oahu_pop <- get_acs(
  geography = "tract",
  state = "HI",
  variables = "B01003_001", # total population
  geometry = TRUE,
  year = 2023
)
```

```{r}
#Download from github "final_data" folder
evac_zones <- st_read(here("final_data/Tsunami_Evacuation_All_Zones.geojson"))
```

```{r}
#Download from github "final_data" folder
traffic <- st_read(here("final_data/Traffic_Analysis_Zones_Oahu.geojson"))
```

```{r}
glimpse(oahu_pop)
```

### Shiny App Code

```{r}
# Choices for the dropdown
available_views <- c("Population Density", "Evacuation Zones", "Density and Evac Zones", "Traffic Data")

ui <- page_sidebar(
  title = "Oahu Tsunami Risk Map",
  sidebar = sidebar(
    selectInput("view_choice", "Choose Data to Display:", choices = available_views)
  ),
  card(
    full_screen = TRUE,
    maplibreOutput("map", height = "100vh"),
    uiOutput("legend_ui")
  )
)

server <- function(input, output, session) {
  filtered_data <- reactive({
    list(
      pop = oahu_pop,
      zones = evac_zones,
      traffic = traffic
    )
  })

  output$map <- renderMaplibre({
    data_list <- filtered_data()
    pop_data <- data_list$pop
    zone_data <- data_list$zones
    traffic_data <- data_list$traffic

    map <- maplibre(style = carto_style("positron"))
    map <- map |> fit_bounds(pop_data, animate = FALSE)

    # Evacuation zone layer
    if (input$view_choice %in% c("Evacuation Zones", "Density and Evac Zones")) {
      map <- map |>
        add_fill_layer(
          id = "evac_zones_layer",
          source = zone_data,
          fill_color = list(
            "match",
            get_column("zone_code"),
            1, "#D96B6B",  # Red for zone 1
            2, "#E5D88F",  # Yellow for zone 2
            3, "#8CCB87",  # Green for zone 3
            "#808080"      # Gray fallback
          ),
          fill_opacity = 0.7,
          popup = "zone_desc",
          tooltip = "zone_type",
          hover_options = list(
            fill_opacity = 0.5
          )
        )
    }

    # Population Density layer
    if (input$view_choice %in% c("Population Density", "Density and Evac Zones")) {
      min_val <- min(pop_data$estimate, na.rm = TRUE)
      max_val <- max(pop_data$estimate, na.rm = TRUE)

      map <- map |>
        add_fill_layer(
          id = "population_layer",
          source = pop_data,
          fill_color = interpolate(
            column = "estimate",
            values = c(min_val, max_val),
            stops = c("#edf8fb", "#3182bd"),
            na_color = "lightgrey"
          ),
          fill_opacity = 0.6,
          popup = "popup",
          tooltip = "estimate",
          hover_options = list(
            fill_color = "yellow",
            fill_opacity = 1
          )
        )
    }

    # Traffic data layer
    if (input$view_choice %in% c("Traffic Data", "All")) {
      min_val <- min(traffic_data$hawaiistgrdsownertaz_oaharea, na.rm = TRUE)
      max_val <- max(traffic_data$hawaiistgrdsownertaz_oaharea, na.rm = TRUE)

      map <- map |>
        add_fill_layer(
          id = "traffic_layer",
          source = traffic_data,
          fill_color = interpolate(
            column = "hawaiistgrdsownertaz_oaharea",
            values = c(min_val, max_val),
            stops = c("#fee5d9", "#de2d26"),
            na_color = "lightgrey"
          ),
          fill_opacity = 0.5,
          tooltip = "hawaiistgrdsownertaz_oaharea",
          hover_options = list(
            fill_opacity = 0.8
          )
        )
    }

    map
  })

  # Dynamic legend
  output$legend_ui <- renderUI({
    if (input$view_choice == "Population Density") {
      tags$div(style = "position:absolute; bottom:10px; left:10px; background:white; padding:8px; border-radius:5px;",
               tags$b("Population Density"),
               tags$div(style = "background:#edf8fb; height:10px; width:60px; display:inline-block; margin-right:5px;"), "Low",
               tags$div(style = "background:#3182bd; height:10px; width:60px; display:inline-block; margin-left:10px;"), "High"
      )
    } else if (input$view_choice == "Evacuation Zones") {
      tags$div(style = "position:absolute; bottom:10px; left:10px; background:white; padding:8px; border-radius:5px;",
               tags$b("Evacuation Zones"), tags$br(),
               tags$div(style = "background:#D96B6B; height:10px; width:20px; display:inline-block;"), " Zone 1", tags$br(),
               tags$div(style = "background:#E5D88F; height:10px; width:20px; display:inline-block;"), " Zone 2", tags$br(),
               tags$div(style = "background:#8CCB87; height:10px; width:20px; display:inline-block;"), " Zone 3"
      )
    } else if (input$view_choice == "Traffic Data") {
      tags$div(style = "position:absolute; bottom:10px; left:10px; background:white; padding:8px; border-radius:5px;",
               tags$b("Traffic Area Size"),
               tags$div(style = "background:#fee5d9; height:10px; width:60px; display:inline-block; margin-right:5px;"), "Small",
               tags$div(style = "background:#de2d26; height:10px; width:60px; display:inline-block; margin-left:10px;"), "Large"
      )
    } else {
      NULL
    }
  })
}

shinyApp(ui, server)

```
